A Refactoring - new attempt

### 2020-08-04 Articles/2020-Refactoring/Source/refactoring-article.tex
More work on the refactoring article

Axioms, definitions and justifications are included in the mathematical section, with the rest of the formal elements. They should be seen as integral part of the framework.

\section{Basic formal structures?}

The starting point of our framework is the \textbf{principle of scientific objectivity} which states \statement{Science is universal, non-contradictory and evidence based}. Every physical theory, then, must give us a set of logically sound statements linked with some notion of experimental verification. The content of the statements, what can be verified and how, will vary from theory to theory, but logical consistency and verifiability must be common to all of them. Therefore we will have two set of axioms: the \textbf{logical consistency axioms} and the \textbf{experimental verifiability axioms}. We start with the first set.

@@

The \textbf{axiom of context} establishes that \textbf{statements} are organized into \textbf{logical contexts}. It is the context that defines their meaning and logical relationships. The definition of a physical object, for example, may change depending on our goal: if we measure the mass of the electron, its value is assumed to be unknown; if we perform particle identification in a detector, the mass of the electron is assumed to be a particular value. Each context has to be such that the meaning of each statement has no ambiguity and therefore it is associated with a single, well-defined truth value. This is, in general, not known: it cannot be proven or deduced. Ultimately, it is the role of experimentation to find it.

@@

On top of these axioms we can define \textbf{falsifiable statements} as the ones that can be disconfirmed experimentally (e.g. their negation is verifiable). We can also also define \textbf{decidable statements} as the ones that can be both verified and falsified experimentally (i.e. their negation is a verifiable statement). One can prove that the finite disjunction and the finite conjunction of decidable statements is also decidable.

@@

\subsection{Domains and possibilities?}

Finite time termination not only imposes a different logical structure, but it also limits the number of statements that can be verified. Even when given an arbitrarily long time, a countable set of statements is the most that can be verified. However, due to the logical relationship between statements, one does not need to test all of them: if \statement{the mass of the photon is less than $10^{-6}$ eV} is experimentally verified, there is no need to test whether \statement{the mass of the photon is greater than $10^{-6}$ eV}.

@@

For each theoretical statement $\tstmt \in \tdomain$, we can define its \textbf{verifiable part} $\ver(\tstmt)$, \textbf{falsifiable part} $\fal(\tstmt)$ and its \textbf{undecidable part} $\und(\tstmt)$. This allows us to precisely characterize the optimal testing strategy in terms of termination. For example, for a decidable statement like $\stmt=$\statement{there are three apples on the table}, the verifiable part will be equal to the statement itself $\ver(\stmt) \equiv \stmt$, the falsifiable part will be equal to its negation cases $\fal(\stmt) \equiv \NOT\stmt$ and the undecidable part will be the contradiction $\und(\stmt) \equiv \contradiction$. The experimental test for $\stmt$ always terminates either successfully or unsuccessfully. Instead, if $\stmt=$\statement{the mass of the photon is less than $10^{-17}$ eV}, we still have $\ver(\stmt) \equiv \stmt$ because it is a verifiable statement, but we will have $\fal(\stmt) \equiv$\statement{the mass of the photon is more than $10^{-17}$ eV} while $\und(\stmt) \equiv$\statement{the mass of the photon is exactly $10^{-17}$ eV}. The issue is that, due to finite precision, if the mass of the photon where exactly $10^{-17}$ eV, our confidence interval would only include the region below that number, so we would never be able to exclude it. So, the test would not terminate in that particular case. Lastly, if $\stmt=$\statement{the mass of the photon is a rational number as expressed in eV}, we have $\ver(\stmt) \equiv \contradiction \equiv \fal(\stmt)$ and $\und(\stmt) \equiv \tautology$. The only way to know whether the statement is true or not would be to know the exact value of the mass, which cannot be achieved in finite time. Therefore the test for $\stmt$ will never terminate. We call such statements \textbf{undecidable}.

@@

There are notable properties conceptual and formal properties that can be proven from such a simple setup. The main results are the following.

\textit{Each statement corresponds to a set of possibilities and logical relationships between statements correspond to set operations.} The idea that any statement in the domain (e.g. \statement{this animal is a mammal}) can be expressed as a disjunction of possibilities (e.g. \statement{this animal is a cat} $\OR$ \statement{this animal is a dog} $\OR$ ...). Each statement, then, can be identified by the set of possibilities that are compatible with it. The conjunction of two statement will be compatible with possibilities that are compatible with both statement. If a statement is narrower than another, then all the possibilities compatible with the first will be compatible with the second. We can therefore reexpress all statement relationships in term of set relationships, as in table \ref{tab:statement_set}.

@@

\textit{The experimental domain induces a natural topology on the possibilities. The theoretical domain induces a $\sigma$-algebra.} As the experimental domain is closed under finite conjunction and countable disjunction, the collections of sets corresponding to verifiable statements is closed under finite intersection and countable unions. As the domain has a countable basis, arbitrary unions become equivalent to countable ones. Therefore the collection of sets associated to verifiable statements is a topology on the set of possibilities. In the same way, the theoretical domain will correspond to a $\sigma$-algebra. Note that this correspond to the actual topologies commonly used. In the standard topology of the real numbers, for example, $(0,1)$ is an open set (a set in the topology) precisely because \statement{the value is between 0 and 1} is a verifiable statement. The singleton $\{1\}$ is a closed set (the complement of a set in the topology) precisely because it is not verifiable due to infinite precision, but it is falsifiable (it can be excluded experimentally). The verifiable part of a statement will correspond to the interior of the set. An undecidable statement will be one with no interior nor exterior. We can map topological concept readily understood in terms of experimental verifiability, as in table \ref{tab:statement_topology}.

@@

\textit{The possibilities are calculated from the verifiable statements.} It is standard in physics to use the infinitely precise values as the starting point, and the finite precision interval as measurement approximations. In our framework it is the opposite: the finite precision verifiable statements are the starting point, and the possibilities are the limits. Technically, one takes the basis of the experimental domain and creates a minterm: a disjuction where each element appears once, negated or not (e.g. $\stmt[e]_1\OR \NOT \stmt[e]_2 \OR ...$). Each minterm that is not a contradiction is a possibility: something that tells us whether each element of the basis is true or note, and therefore any statement in the domain. This approach has various benefits. First, it conceptually maps better to experimental practice. The points of our spaces are the best description possible given a set of experimental techniques. If we expand the experimental techniques (e.g. we can measure the spin of the electron and not just its position), the possibilities are refined. Second, the full descriptions (i.e. the possibilities) and the partial descriptions (i.e. any other statement in the domain) are mathematically the same object. This makes it a lot easier to talk about partial description as they will simply be sub-domains, instead of having to define maps from sets to point and from sets to sets. Third, it makes it easier to understand how the assumptions that give as a particular set of points break down and what are possible alternatives. Assumptions will fail in narrower descriptions, not in the broader ones, so the logical relationships of the broader statements need to be preserved. If we start from the infinitesimal objects, we have to make modifications and then check whether the large scale behavior emerges. Since the framework start by defining broader statements, we are literally making small modifications on our starting points. That is, we keep the large scale structure fixed and see what possible small scale structure can be had.

@@

\textit{The following propositions are all equivalent: all statements in an experimental domain are decidable; all theoretical statements are verifiable; all possibilities are verifiable; the possibilities form a countable basis; the topology is discrete.} This result allows to fully characterize \textbf{decidable domains}, those that contains only decidable statements. The discrete topology one associates with integers corresponds to the fact that we can indeed verify that \statement{there are exactly 3 apples on the table}. We can either verify or falsify each of the cases: all possibilities are decidable. This can only happen if the cases are countable, so if the set of possibilities has the cardinality of the continuum (e.g. the points of a manifold) we must have a qualitatively different space in terms of experimental verifiability.

@@

Having defined experimental domain, we want to define what it means for two domains to have an inference or a causal relationship. The main result is that the two coincide.

Given two domain $\edomain_Y$ and $\edomain_X$, we say there is an \textbf{inference relationship} from the first to the second if by testing a statement in the first we can infer the truth of a statement in the second. Formally, there is a map $\erel : \edomain_Y \to \edomain_X$ such that $\erel(\stmt_Y) \equiv \stmt_Y$. In that case, we say $\edomain_Y$ \textbf{depends} on $\edomain_X$, noted $\edomain_Y \subset \edomain_X$. If $\edomain_Y$ and $\edomain_X$ depend on each other, we say  $\edomain_Y$ is \textbf{equivalent} to $\edomain_X$, noted $\edomain_Y \equiv \edomain_X$.

Given two domain $\edomain_X$ and $\edomain_Y$, we say there is a \textbf{causal relationship} from the first to the second if by knowing which possibility $x \in X$ is true in the first we know which possibility $y \in Y$ is true in the second. Formally, there is a map $f : X \to Y$ such that $x \narrower f(x)$. Note how the direction between the domains

>>>>

Axioms, definitions and justifications are included in the mathematical section, with the rest of the formal elements. The informal elements should be seen as integral part of the framework as we should never be allowed to ``assume'' something that was not fully backed by a scientific explanation.

\section{An overview of the mathematical structures}

The starting point of our framework is the \textbf{principle of scientific objectivity} which states \statement{Science is universal, non-contradictory and evidence based}. Every physical theory, then, must give us a set of logically sound statements together with some notion of experimental verification. While the content of the statements, what can be verified and how, will vary from theory to theory, the rules dictated by logical consistency and experimental verifiability must be always followed. Therefore we will have two set of axioms: the \textbf{logical consistency axioms} and the \textbf{experimental verifiability axioms}. We start with the first set.

@@

The \textbf{axiom of context} establishes that \textbf{statements} are organized into \textbf{logical contexts}. It is the context that defines their meaning and logical relationships. The definition of a physical object, for example, may change depending on our goal: if we measure the mass of the electron, its value is assumed to be unknown; if we perform particle identification in a detector, the mass of the electron is assumed to be a particular value. Each context has to be such that the meaning of each statement has no ambiguity and therefore it is associated with a single, well-defined truth value. This value is, in general, not known, it cannot be proven or deduced. Ultimately, it is the role of experimentation to find it.

@@

On top of these axioms we can define \textbf{falsifiable statements} as the ones that can be disconfirmed experimentally (e.g. their negation is verifiable). We can also also define \textbf{decidable statements} as the ones that can be both verified and falsified experimentally (i.e. their negation is also a verifiable statement). One can prove that the finite disjunction and the finite conjunction of decidable statements is also decidable.

@@

\subsection{Experimental domains and possibilities}

Finite time termination not only imposes a different logical structure, but it also limits the number of statements that can be verified. Even when given an arbitrarily long time, a countable set of statements is the most that can be verified. However, one does not need to test all of statements within a set as some results can be inferred: if \statement{the mass of the photon is less than $10^{-6}$ eV} is experimentally verified, there is no need to test whether \statement{the mass of the photon is greater than $10^{-6}$ eV}.

@@

For each theoretical statement $\tstmt \in \tdomain$, we can define its \textbf{verifiable part} $\ver(\tstmt)$, \textbf{falsifiable part} $\fal(\tstmt)$ and its \textbf{undecidable part} $\und(\tstmt)$. This allows us to precisely characterize the optimal testing strategy in terms of termination. For example, for a decidable statement like $\stmt=$\statement{there are three apples on the table}, the verifiable part will be equal to the statement itself $\ver(\stmt) \equiv \stmt$, the falsifiable part will be equal to its negation cases $\fal(\stmt) \equiv \NOT\stmt$ and the undecidable part will be the contradiction $\und(\stmt) \equiv \contradiction$. The experimental test for $\stmt$ always terminates either successfully or unsuccessfully. Instead, if $\stmt=$\statement{the mass of the photon is less than $10^{-17}$ eV}, we still have $\ver(\stmt) \equiv \stmt$ because it is a verifiable statement, but we will have $\fal(\stmt) \equiv$\statement{the mass of the photon is more than $10^{-17}$ eV} while $\und(\stmt) \equiv$\statement{the mass of the photon is exactly $10^{-17}$ eV}. The issue is that, due to finite precision, if the mass of the photon where exactly $10^{-17}$ eV, our confidence interval would always include the region below that number, so we would never be able to exclude it. So, the test would not terminate in that particular case. Lastly, if $\stmt=$\statement{the mass of the photon is a rational number as expressed in eV}, we have $\ver(\stmt) \equiv \contradiction \equiv \fal(\stmt)$ and $\und(\stmt) \equiv \tautology$. The only way to know whether the statement is true or not would be to know the exact value of the mass, which cannot be achieved in finite time. Therefore the test for $\stmt$ will never terminate. We call such statements \textbf{undecidable}.

@@

There are notable conceptual and formal properties that can be proven from such a simple setup. The main results are the following.

\textit{Each statement in a domain corresponds to a unique set of possibilities and logical relationships between statements correspond to set operations.} The idea that any statement in the domain (e.g. \statement{this animal is a mammal}) can be expressed as a disjunction of possibilities (e.g. \statement{this animal is a cat} $\OR$ \statement{this animal is a dog} $\OR$ ...). Each statement, then, can be identified by the set of possibilities that are compatible with it. The conjunction of two statements will be compatible with possibilities that are compatible with both statements, therefore conjunction corresponds to intersection. If a statement is narrower than another, then all the possibilities compatible with the first will be compatible with the second, therefore the narrowness relationship corresponds to subset relationship. We can re-express all statement relationships in term of set relationships, as in table \ref{tab:statement_set}.

@@

\textit{The experimental domain induces a natural topology on the possibilities. The theoretical domain induces a $\sigma$-algebra.} As the experimental domain is closed under finite conjunction and countable disjunction, the collections of sets corresponding to verifiable statements is closed under finite intersection and countable unions. As the domain has a countable basis, arbitrary unions become equivalent to countable ones. Therefore the collection of sets associated to verifiable statements is a topology on the set of possibilities. In the same way, the theoretical domain will correspond to a $\sigma$-algebra. Note that these correspond to the actual topologies commonly used. In the standard topology of the real numbers, for example, $(0,1)$ is an open set (a set in the topology) precisely because \statement{the value is between 0 and 1} is a verifiable statement. The singleton $\{1\}$ is a closed set (the complement of a set in the topology) precisely because it is not verifiable due to infinite precision, but it is falsifiable (it can be excluded experimentally). The verifiable part of a statement will correspond to the interior of the set. An undecidable statement will be one with no interior nor exterior. Topological concepts can readily be understood in terms of experimental verifiability, as in table \ref{tab:statement_topology}.

@@

\textit{The possibilities are calculated from the verifiable statements.} It is standard in physics to use the infinitely precise values as the starting point, and the finite precision interval as measurement approximations. In our framework it is the opposite: the finite precision verifiable statements are the starting point, and the possibilities are the limits. Technically, one takes the basis of the experimental domain and creates a minterm: a disjuction where each element appears once, negated or not (e.g. $\stmt[e]_1\OR \NOT \stmt[e]_2 \OR ...$). Each minterm that is not a contradiction is a possibility: something that tells us whether each element of the basis is true or not, and therefore any statement in the domain. This approach has various benefits. First, it conceptually maps better to experimental practice. The points of our spaces are the best description possible given a set of experimental techniques. If we expand the experimental techniques (e.g. we can measure the spin of the electron and not just its position), the possibilities are refined. Second, the full descriptions (i.e. the possibilities) and the partial descriptions (i.e. any other statement in the domain) are mathematically the same object. This makes it a lot easier to talk about partial description as they will simply be sub-domains, instead of having to define maps from sets to point and from sets to sets. Third, it makes it easier to understand how the assumptions that give us a particular set of points break down and what are possible alternatives. Assumptions will fail in narrower descriptions, not in the broader ones, so the logical relationships of the broader statements need to be preserved. If we start from the infinitesimal objects, we have to make modifications and then check whether the large scale behavior emerges. Since the framework start by defining broader statements, we are literally making small modifications on our starting points. That is, we keep the large scale structure fixed and see what possible small scale structure can be had.

@@

\textit{The following propositions are all equivalent: all statements in an experimental domain are decidable; all theoretical statements are verifiable; all possibilities are verifiable; the possibilities form a countable basis; the topology is discrete.} This result allows to fully characterize \textbf{decidable domains}, those that contains only decidable statements. The discrete topology one associates with integers corresponds to the fact that statements like \statement{there are exactly 3 apples on the table} are decidable. We can either verify or falsify each of the cases: all possibilities are decidable. This can only happen if the cases are countable, so if the set of possibilities has the cardinality of the continuum (e.g. the points of a manifold) we must have a qualitatively different space in terms of experimental verifiability.

@@

Having defined experimental domain, we proceed to define what it means for two domains to be experimentally related. There are two types of relationships we will consider: inference and causal. The main result is that the two coincide.

Given two domain $\edomain_Y$ and $\edomain_X$, we say there is an \textbf{inference relationship} from the first to the second if by testing a statement in the first we can infer the truth of a statement in the second. Formally, there is a map $\erel : \edomain_Y \to \edomain_X$ such that $\erel(\stmt_Y) \equiv \stmt_Y$. In that case, we say $\edomain_Y$ \textbf{depends} on $\edomain_X$, noted $\edomain_Y \subseteq \edomain_X$. If $\edomain_Y$ and $\edomain_X$ depend on each other, we say  $\edomain_Y$ is \textbf{equivalent} to $\edomain_X$, noted $\edomain_Y \equiv \edomain_X$.

Given two domain $\edomain_X$ and $\edomain_Y$, we say there is a \textbf{causal relationship} from the first to the second if by knowing which possibility $x \in X$ is true in the first we know which possibility $y \in Y$ is true in the second. Formally, there is a map $f : X \to Y$ such that $x \narrower f(x)$. Note how the direction between the domains is different between causal and inference relationship: the inference proceeds from effect to cause. Given these definition, one can prove the following.

\textit{Causal and inference relationships are equivalent.} That is, if there exists a causal relationship between $\edomain_X$ and $\edomain_Y$ then there is an inference relationship between $\edomain_Y$ and $\edomain_X$ and vice-versa. For example, if the temperature of the water determines its density, then by measuring the density we can infer the temperature of the water. We say, then, that there is an \textbf{experimental relationship} between the two domains. Note the relationship is directional: the water density does not determine the temperature because water is most dense at 4 Celsius. In the same way, a measurement on temperature is not in general equivalent to a measurement of density because it can tell us whether we are above or below 4 Celsius, which can't be done with a measurement on density. These are the types of subtleties that a mathematical framework that closely maps to physical concepts helps brings to light.

\textit{Causal relationship are topologically continuous functions.} Given a causal relationship $f : X \to Y$, for every verifiable statement in $\edomain_Y$ we can infer an equivalent statement in $\edomain_X$, which must be verifiable since it is equivalent to a verifiable statement. In terms of the topology, $f^{-1}(U)$ must be an open set if $U$ is open, therefore $f$ must be continuous. This tells us why in physics functions are ``well-behaved''. It is not a matter of convenience: they must be so or they would break experimental verifiability.

\textit{Domain equivalence is isomorphism in the category.} If we were to add additional mathematical structure to our domain, given that the domain equivalence preserves statement equivalence, the additional mathematical structure will be preserves as well. Therefore, if we had a manifold and a differentiable structure, domain equivalence would be a diffeomorphism; if we had a vector space structure, domain equivalence would be a linear transformation and so on. This again tells us when and why these mathematical concepts are important in physics.

\subsection{Combined domains}

Another general operation we can define is domain combination, which takes two or more domains and merges them into one that contains all of them.

Given two domains $\edomain_X$ and $\edomain_Y$ we define the \textbf{combined domain} $\edomain_X \times \edomain_Y$ as the one that contains all the verifiable statements that can be constructed from both $\edomain_X$ and $\edomain_Y$. For example, if one domain describes the shape of an object and the other its color, the combined domain will contain descriptions of both. Formally, $\edomain_X \times \edomain_Y$ is the closure under finite conjunction and countable disjunction. This definition can be extended to the case of countably many domains.

The goal is to understand how the possibilities depend on the type of relationship we have between the domains. We therefore introduce the following definitions. Two domains are said to be \textbf{independent} if taking a statement from each domain always gives a set of independent statements. Formally, $\stmt_X \indep \stmt_Y$ for all $\stmt_X \in \edomain_X$ and $\stmt_Y \in \edomain_Y$. Two domains are said to be \textbf{incompatible} if all statements from one are incompatible with all statements from the other, with the exception of tautologies. Formally, $\stmt_X \ncomp \stmt_Y$ for all $\stmt_X \in \edomain_X \setminus \{\tautology\}$ and $\stmt_Y \in \edomain_Y \setminus \{\tautology\}$. Now a list of consequences for these definitions.

\textit{The set of possibilities of the combined domain is a subset of the scalar product of the possibilities of the individual domains.} Intuitively, a complete description of the individual domains is equivalent to the complete description of both domain combined. Not all combinations, though, may be possible because there may be a dependence between the two domains.\footnote{It is an open question what are the possible subsets of the scalar product are allowed for a combined domain.}

\textit{If $\edomain_Y$ depends on $\edomain_X$, then the combined domained is $\edomain_X$.} Intuitively, if $\edomain_Y$ depends on $\edomain_X$ (e.g. the domain for water density depends on the one for water temperature), then $\edomain_X$ already contains all statements in $\edomain_Y$ and all its possibilities, so nothing is added by the combination.

\textit{If $\edomain_X$ is independent of $\edomain_Y$, then the combined possibilities coincides with the scalar product.} If the domains are independent (e.g. vertical position and horizontal position), so will be their possibilities. Each possibility from the first can be combined with any possibility of the second, so all combinations are allowed. The natural topology will correspond to the product topology even in the case of countably many domains.\footnote{Note that it will naturally correspond to the product topology, and not the box topology, because of closure under finite conjunction.}

\textit{If $\edomain_X$ is incompatible to $\edomain_Y$, then the combined possibilities coincides with the disjoint union of the possibilities.} If the domains are incompatible (e.g. descriptions for a plant and descriptions for an animal), the possibilities will also describe incompatible cases. So, conceptually, the each combined possibility will be a possibility from one of the two domains. The details are a bit more involved and they are available in the appendix.
